Do larger catchments respond different to forest cover change? Re-analysing a global data set.

### 2021-09-23 paper/Forest_and_Water/Forest_and_Water.Rmd
update paper with first analysis

Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover. However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f. @jackson2005) and possibly the climate (as indicated by either Pa and E0 or latitude and longitude @filoso2017; @zhou2015).

However, there could be further confounding factors, which are eluded to by @filoso2017:  
- the type of analysis, i.e. paired catchment studies, modelling, time series analysis etc.  
- the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.

@@

$$\tag{1}
\begin{aligned}
\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + {climate~type} + \varepsilon
\end{aligned}$$

@@

$$\tag{2}
\begin{aligned}
log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + {climate~type} + \varepsilon
\end{aligned}$$

@@

>>>>

<!----
The starting point consists of the two tables in the supplementary data from Zhang et al. (2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2). We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
--->
```{r, echo = F, message = F}
require(tidyverse)
require(lubridate)

Zhang_small <- read_csv("data/TablesZhangetal2017_small.csv")
Zhang_large <- read_csv("data/TablesZhangetal2017_large.csv")
```

<!--- ### change names of columns

the package`mgcv` does not seem to be happy with complex names, so probably better to make all names simple --->

```{r, echo = F}
names(Zhang_small)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_large)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_small)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_large)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_small)[9:10] <- c("Precip_data_type", "Assessment_technique")
names(Zhang_large)[9:10] <- c("Precip_data_type", "Assessment_technique")
```

<!-- combine both tables -->

```{r, echo = F}
Zhang_small$From <- as.numeric(Zhang_small$From)
Zhang_small$To <- as.numeric(Zhang_small$To)
Zhang_all <- bind_rows(Zhang_large,Zhang_small)
```

Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover. However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f. @jackson2005) and possibly the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).

However, there could be further confounding factors, which are eluded to by @filoso2017:

- the type of analysis, i.e. paired catchment studies, modelling, time series analysis etc.

- the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.

@@

$$\tag{1}
\begin{aligned}
\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + Latitude + Longitude + \varepsilon
\end{aligned}$$

@@

$$\tag{2}
\begin{aligned}
log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + Latitude + Longitude + \varepsilon
\end{aligned}$$

@@

Results
===========

<!-- analysis -->
```{r model2, echo = F}
require(mgcv)
linear_model_all_log <- gam(log10(DeltaQf_perc/100+1) ~ DeltaF_perc + Area_km2 + Pa_mm, data = Zhang_all)
summary(linear_model_all_log)
gam.check(linear_model_all_log)
```

### 2021-09-23 paper/Forest_and_Water/Forest_and_Water.tex
update paper with first analysis

There has been an long and on-going discussion in the hydrologcal
literature around the impact of forests on streamflow (Andréassian,
2004; Brown et al., 2013, 2005; Jackson et al., 2005; Zhang et al.,
2017). The historic work highlights provides a general consensus that if
forest areas increase, streamflow decreases and vice-versa. The most
dramatic result in relation to this, is Figure 5 in Zhang et al. (2011)
indicating (or Australian watersheds) a 100\% decrease in stream flow
for watersheds with 100\% forest cover. However, on the other end of the
spectrum, in a series of French watersheds (Cosandey et al., 2005),
there was no change in streamflow characteristics in 2 of the three
watersheds studied in relation to deforestation.

There have been several review papers aiming to summarize different
studies across the globe, in relation to paired watershed studies (Bosch
and Hewlett, 1982; Brown et al., 2005) and more generally (Jackson et
al., 2005; Zhang et al., 2017). These studies are aiming to generalize
the individual findings and to identify if there are global trends or
relationships that can be developed. The most recent review (Zhang et
al., 2017) developed an impressive database of watershed studies in
relation to changes in streamflow due to changes in forest cover based
on a global data set. This dataset, which covers over 250 studies are
described in terms of the change in streamflow as a result of the change
in forest cover, where studies related to both forestation (increase in
forest cover) and deforestation (decrease in forest cover) were
included.

The conclusions of the paper (Zhang et al., 2017) suggest that there is
a distinct difference in the change in flow as a result of forestation
or deforestation between small watersheds, defined as \textless{} 1000
km\textsuperscript{2} and large watersheds \textgreater{} 1000
km\textsuperscript{2}. While for small watersheds there was no real
change in runoff with changes in cover, for large watersheds there was a
clear trend showing a decrease in runoff with and increase in forest
cover. Their main conclusion was that the response in annual runoff to
forest cover was scale dependent and appeared to be more sensitive to
forest cover change in water limited watersheds relative to energy
limited watershed (Zhang et al., 2017).

Encouraged by the work presented by Zhang et al. (2017) and the
fantastic database of studies presented by these authors, we believe we
can add to the discussion by presenting further analysis of the data and
by adding further watersheds and enhancements to the data base.

In particular, the main method in the work by Zhang et al. (2017) is
using simple linear regression. And the main assumption is that the
threshold at 1000 km\textsuperscript{2} is a distinct separation between
``small'' and ``large'' watersheds. Given the fantastic data set
collected, the analysis can be easily expanded to look at interactions
between the terms and to test the assumption of a distinct threshold at
1000 km\textsuperscript{2}.

In particular, the objective of this paper is to 1) enhance the data set
from Zhang et al. (2017) with further watersheds and spatial coordinates
and 2) to analyse the possibility of non-linear and partial effects of
the different factors and variables in the data base using generalised
linear (GLM) and generalised additive models (GAM Wood (2006)). Finally
we hope to point to further research that can expand our work and that
outlined Zhang et al. (2017) to better understand the impact of forest
cover change on streamflow.

@@

\hypertarget{the-original-data-set}{%
\subsection{The original data set}\label{the-original-data-set}}

@@

While the Zhang et al. (2017) use the dryness index in their analysis,
potential or reference evapotranspiration is not part of the data set.
We combined the tables for small (\textless{} 1000
km\textsuperscript{2}) and large (\textgreater= 1000
km\textsuperscript{2}) watreshed data sets in our analysis. Some small
naming errors and citations for some of the data sets for some of the
small watersheds were fixed as we were familiar with the studies. But
overall the original data set was not changed.

@@

To enhance the existing data set, this study added additional variables.
The first variables added were the latitude and longitude for the center
of the watershed as an approximation of its spatial location. Using this
information annual average reference evapotranspiration (E0) was
extracted from \textbf{XXXXX} if value of E0 were not available from the
original papers. For large water watersheds, this value, similar to
annual average rainfall, is an approximation of the climate at the
location.

The length of the study can be a variable influencing the change in flow
(e.g.~Jackson et al., 2005) and therefore, the length of the different
studies was extracted from the references provided by Zhang et al.
(2017).

@@

The final column in the improved data set is a ``notes'' column, which
is not further used in the analysis, but gives context to some of the
data for future research.

@@

To estimate how the change in streamflow is affected by the change in
forest cover while considering the effects of the other variables, we
applied generalised additive modelling (GAM) (Wood, 2006). In particular
we fitted the following initial model to replicate the variables in the
analysis from Zhang et al. (2017):

\[\tag{1}
\begin{aligned}
\Delta \%Q \sim ~&\Delta \%forest + s(Pa) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

However, the overall skewed distribution of the predictant
(\(\Delta \%Q\)) is problematic, and this results in a skewed
distribution of the GAM model residuals. As a result we transformed
\(\Delta \%Q\) and \(\Delta \%forest\) back to fractions (0 - 1) and log
transformed using \(log10(x + 1)\), where \(x\) is either \(\Delta Q\)
or \(\Delta forest\). This means that the model residuals are
\(\sim N(0,\sigma^2)\) results in the following equation:

\[\tag{2}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

In this model, the assumption is that all continuous variables (such as
Pa) can have a linear or non-linear relationship with
\(log10(\Delta Q)\). This means that a smooth function \(s()\) is
applied to the variable. To restrict the smoothness of the fit, the
smoothness factor \(k\) is restricted to a value of 3 (Wood, 2006). This
restriction was applied to smooth variables throughout this paper and we
have dropped this from the subsequent equations.

For the model in equation 2, we only used the data from Zhang et al.
(2017) to make sure that the additional watersheds added to the data set
did not influence the analysis. Given that in Zhang et al. (2017),
dryness (\(\frac{E0}{Pa}\)) is used to look at variations in the change
in flow, we also fitted the following model:

\[\tag{3}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

@@

\[\tag{4}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
\end{aligned}\]

\[\tag{5}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
\end{aligned}\]

@@

@@

>>>>

There has been an long and on-going discussion in the hydrologcal
literature around the impact of forests on streamflow (Andréassian,
2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al.,
2005; Zhang et al., 2017). The historic work highlights a general
consensus that if forest areas increase, streamflow decreases and
vice-versa. The most dramatic result in relation to this, is Figure 5 in
Zhang et al. (2011) indicating (for Australian watersheds) a 100\%
decrease in stream flow for watersheds with 100\% forest cover. However,
on the other end of the spectrum, in a series of French watersheds
(Cosandey et al., 2005), there was no change in streamflow
characteristics in 2 of the three watersheds studied in relation to
deforestation.

Several review papers have summarized different studies across the
globe, in relation to paired watershed studies (Bosch and Hewlett, 1982;
Brown et al., 2005), related to reforestation in particular (Filoso et
al., 2017), and more generally (Jackson et al., 2005; Zhang et al.,
2017). These studies aim to generalize the individual findings and to
identify if there are global trends or relationships that can be
developed. The most recent reviews (Filoso et al., 2017; Zhang et al.,
2017) developed an impressive global database of watershed studies in
relation to changes in streamflow due to changes in forest cover. The
Zhang et al. (2017) dataset, which covers over 250 studies, is described
in terms of the change in streamflow as a result of the change in forest
cover, where studies related to both forestation (increase in forest
cover) and deforestation (decrease in forest cover) were included. In
contrast, the paper by Filoso et al. (2017) focused primarily on
reforestation, and covered an equally impressive database of 167 studies
using a systematic review. In this case the collected data is mostly
coded as count data and only a subset of 37 studies was analysed for
actual water yield change.

The conclusions of the first paper (Zhang et al., 2017) suggest that
there is a distinct difference in the change in flow as a result of
forestation or deforestation between small watersheds, defined as
\textless{} 1000 km\textsuperscript{2} and large watersheds
\textgreater{} 1000 km\textsuperscript{2}. While for small watersheds
there was no real change in runoff with changes in cover, for large
watersheds there was a clear trend showing a decrease in runoff with and
increase in forest cover. Their main conclusion was that the response in
annual runoff to forest cover was scale dependent and appeared to be
more sensitive to forest cover change in water limited watersheds
relative to energy limited watershed (Zhang et al., 2017).

The second study (Filoso et al., 2017) was a systematic review which
classified the historical research and highlighted gaps in the spatial
distribution, the types of studies and the types of analysis. Their main
conclusion was also that reforestation decreases streamflow, but that
there were many interacting factors. For a subset of quantitative data
(37) they showed a relationship between catchment size and decline in
streamflow.

A final summary paper that includes much of the same data as Zhang et
al. (2017) and Filoso et al. (2017) is Zhou et al. (2015), which has one
author in common with Zhang et al. (2017). However, this paper aims to
explain the variation in the data using the Fuh model, and in particular
aims to link the variation in the observed data to variations in the
exponent \emph{m} in the model. A key observation is that in drier
environments, the effects of deforestation are much greater than in
wetter environments, which is also suggested by Figure 4 in Zhang et al.
(2017).

Encouraged by the work presented by Zhang et al. (2017) and Filoso et
al. (2017) and the fantastic database of studies presented by these
authors, we believe we can add to the discussion. In this paper, the aim
is to develop further analysis of the collected data and expanding and
combining the two data sets to provide further depth.

In particular, the main method in the work by Zhang et al. (2017) is
using simple linear regression, and in Filoso et al. (2017) the focus is
mainly on classification. As Zhang et al. (2017) points out, the main
assumption in their work is that the threshold at 1000
km\textsuperscript{2} is a distinct separation between ``small'' and
``large'' watersheds, but the subset of data in Filoso et al. (2017)
does not appear to support this. And while te work Filoso et al. (2017)
provides important insights in study types, analysis types and broad
classification, there is limited quantification of actual impact. This
is because the work had a strict criterion to select quantitative
studies. However, given the fantastic data sets collected, the analyses
can be easily expanded to look at interactions between the terms and to
test the assumption of a distinct threshold at 1000
km\textsuperscript{2}.

As a result the objective of this paper is to 1) enhance the data set
from Zhang et al. (2017) with further watersheds (such as from Filoso et
al. (2017)) and spatial coordinates and 2) to analyse the possibility of
non-linear, interactions and partial effects of the different factors
and variables in the data using generalised linear (GLM) and generalised
additive models (GAM Wood (2006)). Finally, this work aims to point to
further research that can expand this area of work, based on the
collected data, to better understand the impact of forest cover change
on streamflow.

@@

\hypertarget{the-original-data-sets}{%
\subsection{The original data sets}\label{the-original-data-sets}}

@@

While Zhang et al. (2017) use the dryness index in their analysis,
potential or reference evapotranspiration was not originally included as
part of the published data set. We combined the tables for small
(\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000
km\textsuperscript{2}) watershed data sets in our analysis.

@@

To enhance the existing data set, this study added additional variables
and cross-checked the studies with the data set from Filoso et al.
(2017). In particular, we focussed on the 37 data points included in the
quantitative analysis in Filoso et al. (2017).

In addition, additional variables added were the latitude and longitude
for the center of the watershed as an approximation of its spatial
location. Using this information annual average reference
evapotranspiration (E0) was extracted from \textbf{XXXXX} if a value of
E0 was not available from the original papers. For large watersheds,
this value, similar to annual average rainfall, is only an approximation
of the climate at the location.

The length of the study can be a variable influencing the change in flow
(e.g.~Jackson et al., 2005) and therefore, the length, starting data and
completion date of the different studies was extracted from the
references provided by Zhang et al. (2017).

@@

The final column in the improved data set is a ``notes'' column, which
is not further used in the analysis, but gives context to some of the
data for future research and highlights some of the discrepancies that
we found between the original papers and the data in the tables from
Zhang et al. (2017).

@@

Building on the analyses by Zhang et al. (2017) and Filoso et al.
(2017), and combining their conclusions, the main hypothesis to test is
that the change in streamflow is impacted by the change in forest cover.
However, this change is clearly modulated by the area under
consideration (affecting the length of the flowpaths Zhou et al.
(2015)), the length of the study (c.f. Jackson et al. (2005)) and
possibly the climate (as indicated by either Pa and E0 or latitude and
longitude Filoso et al. (2017); Zhou et al. (2015)).

However, there could be further confounding factors, which are eluded to
by Filoso et al. (2017):\\
- the type of analysis, i.e.~paired catchment studies, modelling, time
series analysis etc.\\
- the age of the study, assuming that historical studies might not have
had the ability to measure at the accuracy that currently is available
to researchers, or that more careful historical attention to detail in
field studies might have been lost more recently due to reductions in
research investment.

To estimate how the change in streamflow is affected by the change in
forest cover while considering the effects of the other variables, we
applied generalised additive modelling (GAM) (Wood, 2006).

The first model applied in this analysis is based on the main hypothesis
outlined above, can the change in streamflow be predicted from the
change in forest cover, modulated by area, the length of the study and
the climate.

\[\tag{1}
\begin{aligned}
\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + {climate~type} + \varepsilon
\end{aligned}\]

However, the overall skewed distribution of the predictant
(\(\Delta \%Q\)) is problematic, and this results in a skewed
distribution of the GAM model residuals, which violates the linear model
assumptions. As a result we transformed \(\Delta \%Q\) and
\(\Delta \%forest\) back to fractions (0 - 1) and log transformed using
\(log10(x + 1)\), where \(x\) is either \(\Delta Q\) or
\(\Delta forest\). This means that the model residuals approximate
\(\sim N(0,\sigma^2)\) and this results in the following equation:

\[\tag{2}
\begin{aligned}
log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + {climate~type} + \varepsilon
\end{aligned}\]

A second model included all the variables in the analysis from Zhang et
al. (2017) in one model:

\[\tag{3}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

In this model, no direct interactions are assumed, and the assumption is
that all continuous variables (such as Pa) can have a linear or
non-linear relationship with \(log10(\Delta Q)\). This means that a
smooth function \(s()\) is applied to the variable. To restrict the
smoothness of the fit, the smoothness factor \(k\) is restricted to a
value of 3 (Wood, 2006). This restriction was applied to smooth
variables throughout this paper and we have dropped this from the
notation in subsequent equations.

For the model in equation 3, we only used the data from Zhang et al.
(2017) to make sure that the additional watersheds added to the data set
did not influence the analysis. Given that in Zhang et al. (2017),
dryness (\(\frac{E0}{Pa}\)) is used to look at variations in the change
in flow, we also fitted the following model:

\[\tag{4}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

@@

\[\tag{5}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(begin_{year}) + s(length_{study}) + {forest~type} + \\ & {climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

\[\tag{6}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + \\ 
& s(Latitude) + s(Longitude) + s(begin_{year}) + s(length_{study}) + \\ 
& {forest~type} + {climate~type} + {assessment~type} + \\  
& {hydrologic~regime} + \varepsilon
\end{aligned}\]

@@

\leavevmode\hypertarget{ref-filoso2017}{}%
Filoso, S., Bezerra, M.O., Weiss, K.C.B., Palmer, M.A., 2017. Impacts of
forest restoration on water yield: A systematic review. PLOS ONE 12,
e0183210.
doi:\href{https://doi.org/10.1371/journal.pone.0183210}{10.1371/journal.pone.0183210}

@@

\leavevmode\hypertarget{ref-zhou2015}{}%
Zhou, G., Wei, X., Chen, X., Zhou, P., Liu, X., Xiao, Y., Sun, G.,
Scott, D.F., Zhou, S., Han, L., Su, Y., 2015. Global pattern for the
effect of climate and land cover on water yield. Nature Communications
6, 5918.
doi:\href{https://doi.org/10.1038/ncomms6918}{10.1038/ncomms6918}
